---
author: "Richard D. Wilkinson"
date: "03/11/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of NOAA storm event data

## Synopsis

This report analyses data from the NOAA storm database in order to provide tentative answers to two questions: (1) which types of storm events are most harmful to population health and (2) which types of storm events have the greatest economic consequences? After noting some of the datasets limitations (crop damage estimates only from 1993 onward, discrepances in damage estimates, incoherent storm types), this report shows on the one hand that "tornados", "excessive heat" and "TSTM wind" were the worst storm types in terms of population health impacts. On the other hand, "floods", "tornados" and "storm surges" were the three most catastrophic storm types in terms of estimated damage costs (property and crop damage).


## Data processing
### Pre-processing
Check if the dataset is already available in the adequate folder, otherwise create this and download it.
```{r}
if (!dir.exists(file.path("data"))) {
  dir.create(file.path("data"))
} 

if (!file.exists("data/StormData.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/StormData.csv.bz2")
}
```

Load the data into the environment.
```{r, cache = TRUE}
stormData <- read.csv("data/StormData.csv.bz2", header = TRUE, sep = ",")
```

Get an idea of how the dataset is structured.
```{r}
names(stormData)
head(stormData)
```
Keep the variables of interest for the analysis and make sure they are of the correct type (switch the event type variable to a factor).
```{r}
newStormData <- stormData[, c("REFNUM", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP",  "EVTYPE", "BGN_DATE", "REMARKS")]

for (i in 1:dim(newStormData)[2]) {
  str(newStormData[i])
}

stormData$EVTYPE <- as.factor(stormData$EVTYPE)
```

#### Do we have event types for every year in the dataset?
The date is not in a format that is recognisable by R. Let's strip the year from the BGN_DATE variable.

```{r}
head(newStormData)

library(tidyverse)

newStormData$newDate <- str_split_fixed(newStormData$BGN_DATE, " ", 2)[, 1]
newStormData$newYear <- str_split_fixed(newStormData$newDate, "/", 3)[, 3]

missingEventTypes <- newStormData %>%
  group_by(newYear) %>%
  summarise(SumEventTypes = sum(is.na(EVTYPE)))

head(missingEventTypes)
```
There are no years with missing event types.

#### What level of consistency was achieved in the categorisation of the events?
```{r}
head(unique(newStormData$EVTYPE))
length(unique(newStormData$EVTYPE))
```

There are 985 different types of storm events. Some of these categories could be collapsed (e.g. "FLASH FLOOD LANDSLIDES" & "FLASH FLOOD/LANDSLIDE"). Others are not actually storm types (e.g. "Summary of March 14"). It is assumed that more catastrophic events will have been coded more consistently, so event categories are not collapsed here. Nevertheless, records with event types labelled as "summar*" are removed from the dataset.
```{r}
newStormData$EVTYPE <- toupper(newStormData$EVTYPE)

newStormData$EVTYPE2 <- newStormData$EVTYPE
newStormData$EVTYPE2[str_detect(newStormData$EVTYPE, "SUMMAR") == TRUE] <- NA
sum(is.na(newStormData$EVTYPE2))

newStormData <- newStormData[!is.na(newStormData$EVTYPE2),]
head(newStormData)
```

#### Distribution of the estimates of injured and fatalities
```{r}
summary(newStormData$INJURIES)
summary(newStormData$FATALITIES)
```

#### Preparing the damage estimates
Damage is measured bidimensionally in the dataset, through an estimate of the cost of damage to property, as well as of the cost of damage to crops. Both estimates, as printed in the NOAA reports, comprise a base estimate and a multiplier. However, apart from h = hundred, k = thousand, m = million and b = billion, the labels are unclear ("", "?", "0", "+").
```{r}
unique(newStormData$PROPDMGEXP)
unique(newStormData$CROPDMGEXP)
```

Let's make all the mutipliers uppercase to collapse some of the categories.
```{r}
newStormData$PROPDMGEXP <- toupper(newStormData$PROPDMGEXP)
newStormData$CROPDMGEXP <- toupper(newStormData$CROPDMGEXP)
```

Can the meaning of the numbered values be guessed qualitatively from the remarks? Let's compare some of these.
```{r}
m0 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "0" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)
m1 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "1" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)
m2 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "2" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)
m3 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "3" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)
m4 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "4" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)
m5 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "5" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)
m6 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "6" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)
m7 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "7" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)
m8 <- head(newStormData$REMARKS[newStormData$PROPDMGEXP == "8" & newStormData$REMARKS != "" & newStormData$REMARKS != "  "], 2)

listRemarks <- list(M0 = m0, M1 = m1, M2 = m2, M3 = m3, M4 = m4, 
                    M5 = m5, M6 = m6, M7 = m7, M8 = m8)

listRemarks
```
There does seem to be some level of gradation in these descriptions. Hence, it seems as though there are two competing measurement systems: the h, k, m, b system and a numeric scale ranging from 0 to 8.

Which is the most frequently used?

```{r}
newStormData$PROPDMGEXP[newStormData$PROPDMGEXP == ""] <- "NULL"
newStormData$CROPDMGEXP[newStormData$CROPDMGEXP == ""] <- "NULL"

tallySystems <- newStormData %>%
  group_by(PROPDMGEXP) %>%
  tally()
  
tallySystems
```

It seems the letter system was used more frequently than the numeric scale (at least for the property damage). Let's remove the records with numeric damage multipliers.
```{r}
newStormData$numericDamage <- str_detect(newStormData$PROPDMGEXP, "[:digit:]")
table(newStormData$numericDamage)
newStormData$numericDamage2 <- str_detect(newStormData$CROPDMGEXP, "[:digit:]")
table(newStormData$numericDamage2)

newStormData <- subset(newStormData, newStormData$numericDamage == FALSE)
newStormData <- subset(newStormData, newStormData$numericDamage2 == FALSE)

dim(newStormData)
```

Let's also remove the records with "-", "+", and "?".
```{r}
newStormData$numericDamage3 <- str_detect(newStormData$PROPDMGEXP, "\\W")
table(newStormData$numericDamage3)
newStormData$numericDamage4 <- str_detect(newStormData$CROPDMGEXP, "\\W")
table(newStormData$numericDamage4)

newStormData <- subset(newStormData, newStormData$numericDamage3 == FALSE)
newStormData <- subset(newStormData, newStormData$numericDamage4 == FALSE)

dim(newStormData)
```

Let's tally the records by multiplier again.
```{r}
tallySystems <- newStormData %>%
  group_by(PROPDMGEXP) %>%
  tally()
  
tallySystems

tallySystems2 <- newStormData %>%
  group_by(CROPDMGEXP) %>%
  tally()
  
tallySystems2
```

That's looking cleaner!

Now, let's calculate the damage estimate with these.
```{r}
newStormData$CROPDMGEXP.num[newStormData$CROPDMGEXP == "B"] <- 1000000000 # Billion
newStormData$CROPDMGEXP.num[newStormData$CROPDMGEXP == "M"] <- 1000000 # Million
newStormData$CROPDMGEXP.num[newStormData$CROPDMGEXP == "K"] <- 1000 # Thousand
newStormData$CROPDMGEXP.num[newStormData$CROPDMGEXP == "H"] <- 100 # Hundred
newStormData$CROPDMGEXP.num[newStormData$CROPDMGEXP == "NULL"] <- 1 # 1

newStormData$cropDmgEst <- newStormData$CROPDMG * newStormData$CROPDMGEXP.num
summary(newStormData$cropDmgEst)

newStormData$PROPDMGEXP.num[newStormData$PROPDMGEXP == "B"] <- 1000000000 # Billion
newStormData$PROPDMGEXP.num[newStormData$PROPDMGEXP == "M"] <- 1000000 # Million
newStormData$PROPDMGEXP.num[newStormData$PROPDMGEXP == "K"] <- 1000 # Thousand
newStormData$PROPDMGEXP.num[newStormData$PROPDMGEXP == "H"] <- 100 # Hundred
newStormData$PROPDMGEXP.num[newStormData$PROPDMGEXP == "NULL"] <- 1 # 1

newStormData$propDmgEst <- newStormData$PROPDMG * newStormData$PROPDMGEXP.num
summary(newStormData$propDmgEst)
```

#### How has crop and property damage evolved over time?

The following figure is a multiple line chart of the average cost by year of damage to property and crops over the available time period.
```{r}
averages <- newStormData %>%
  group_by(newYear) %>%
  summarise_at(c("propDmgEst", "cropDmgEst"), mean, na.rm = TRUE) %>%
  gather(key = "variable", value = "value", -newYear)

averages$newDate <- paste(averages$newYear, "1", "1", sep = "/")
averages$newDate <- as.Date(averages$newDate)

library(ggplot2)
ggplot(data = averages, aes(x = newDate, y = value, group = variable)) +
  geom_line(aes(color = variable)) + 
  labs(title = "Multiple line chart of the mean crop and property damage estimates by year",
       caption = "Data: NOAA Storm Database.",
       x = "Year",
       y = "Estimated cost",
       fill = "Legend:") +
  scale_color_discrete(name = "Legend:", labels = c("Crop damage", "Property damage"))
  

```
Here, it becomes apparent that there are no crop-damage estimates before 1993. This is a noteworthy limitation of the dataset.

### Data analysis
#### Q1: Which events are most harmful to population health?

Let's calculate the total number of injured and fatalities by event type.
```{r}
highestInjuries <- newStormData %>%
  group_by(EVTYPE2) %>%
  summarise_at("INJURIES", sum, na.rm = TRUE) %>%
  slice_max(INJURIES, n = 10) %>%
  mutate(Pos = dense_rank(desc(INJURIES)))

names(highestInjuries)[2] <- "Num"
highestInjuries$type <- "injuries"

highestFatalities <- newStormData %>%
  group_by(EVTYPE2) %>%
  summarise_at("FATALITIES", sum, na.rm = TRUE) %>%
  slice_max(FATALITIES, n = 10) %>%
  mutate(Pos = dense_rank(desc(FATALITIES)))

names(highestFatalities)[2] <- "Num"
highestFatalities$type <- "fatalities"
```

#### Q2: Which events have the greatest economic consequences?
```{r}
highestPropDmg <- newStormData %>%
  group_by(EVTYPE2) %>%
  summarise_at("propDmgEst", sum, na.rm = TRUE) %>%
  slice_max(propDmgEst, n = 10) %>%
  mutate(Pos = dense_rank(desc(propDmgEst)))

names(highestPropDmg)[2] <- "Num"
highestPropDmg$type <- "propDmg"

highestCropDmg <- newStormData %>%
  group_by(EVTYPE2) %>%
  summarise_at("cropDmgEst", sum, na.rm = TRUE) %>%
  slice_max(cropDmgEst, n = 10) %>%
  mutate(Pos = dense_rank(desc(cropDmgEst)))

names(highestCropDmg)[2] <- "Num"
highestCropDmg$type <- "cropDmg"
```




## Results
### Q1: Which events are most harmful to population health?

The ten worst event types, in terms of both numbered of injured and fatalities, are the following:
```{r}
mergedRank <- data.frame(Rank = highestFatalities$Pos,
                         WorstEventsFatalities = highestFatalities$EVTYPE2,
                         NumFatalities = highestFatalities$Num,
                         WorstEventsInjured = highestInjuries$EVTYPE2,
                         NumInjured = highestInjuries$Num)

mergedRank
```

The following figure is a stacked bar chart of the estimated number of fatalities and people injured due to the ten most high-impact event types.
```{r}
mergedRank2 <- rbind(highestInjuries, highestFatalities)

ggplot(data = mergedRank2, aes(x = reorder(factor(EVTYPE2), -Num), y = Num, fill = type)) +
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  labs(title = "Stacked bar chart of estimated health impacts by event type",
       caption = "Data: NOAA Storm Database.",
       x = "Event type",
       y = "Estimated nb of injured/fatalities",
       fill = "Legend:") +
  scale_fill_discrete(name = "Legend:", labels = c("Fatalities", "Injured"))
```

### Q2: Which events have the greatest economic consequences?
On the other hand, the ten worst event types, in terms of property and crop damage, are the following:
```{r}
mergedRank3 <- data.frame(Rank = highestPropDmg$Pos,
                         WorstEventsPropertyDmg = highestPropDmg$EVTYPE2,
                         CostPropDmg = highestPropDmg$Num,
                         WorstEventsCropDmg = highestCropDmg$EVTYPE2,
                         CostCropDmg = highestCropDmg$Num)

mergedRank3
```

The following figure is a stacked bar chart of the estimated cost of the property and crop damage caused by the ten most high-impact event types.
```{r}
mergedRank4 <- rbind(highestPropDmg, highestCropDmg)

ggplot(data = mergedRank4, aes(x = reorder(EVTYPE2, -Num), y = Num, fill = type)) +
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 75, hjust = 1)) + 
  labs(title = "Stacked bar chart of estimated cost of damage by event type",
       caption = "Data: NOAA Storm Database.",
       x = "Event type",
       y = "Estimated cost of damage") +
  scale_fill_discrete(name = "Legend:", labels = c("Crop damage", "Property damage"))
```

