---
title: "PeerAssignment2"
author: "Richard D. Wilkinson"
date: "03/11/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis


# Data processing
## Pre-processing
Check if the dataset is already available in the adequate folder, otherwise create this and download it.
```{r}
if (!dir.exists(file.path("data"))) {
  dir.create(file.path("data"))
} 

if (!file.exists("data/StormData.csv.bz2")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/StormData.csv.bz2")
}
```

Load the data into the environment.
```{r}
stormData <- read.csv("data/StormData.csv.bz2", header = TRUE, sep = ",")
head(stormData)
```

Get an idea of how the dataset is structured.
```{r}
names(stormData)
head(stormData)
```
Keep the variables of interest for the analysis and make sure they are of the correct type.
```{r}
newStormData <- stormData[, c("REFNUM", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP",  "EVTYPE")]

for (i in 1:dim(newStormData)[2]) {
  str(newStormData[i])
}

stormData$EVTYPE <- as.factor(stormData$EVTYPE)
```

There are 985 different types of storm events. Some of these categories could be collapsed (e.g. "FLASH FLOOD LANDSLIDES" & "FLASH FLOOD/LANDSLIDE"). Others are not actually storm types (e.g. "Summary of March 14"). However, this is not performed here. It is assumed that there will be more consistency in the naming of the most frequent and catastrophic event types.

## Analysis

```{r}

```



# Results